Reduced diversity and increased specialization of Alaska salmon fishermen:

### 2016-10-24 salmon/paper/salmon_paper.Rmd
salmon paper edits from Alan

---
title: 'Reduced diversity and increased specialization of Alaska salmon fishermen:
  effects on variation in individual revenues'
author: "Eric, Sean, Ole, Rich, Milo, Jordan, Jen, others"
date: ''
output:
  pdf_document:
    fig_caption: yes
  html_document: default
  word_document: default
bibliography: refs.bib
---

@@

On a large scale, individual fishers may experience positive outcomes from diversifying their catches; examples include reduced year-to-year variability in annual revenue, and increases in average revenue. Fishers can increase their species diversity by owning a permit that allows multiple species to be targeted, or by owning multiple permits. External forces such as management regulations (limited entry,  catch shares) or market forces (permit prices) may disincentive individuals to hold multiple permits. Thus, individuals with a single permit may increase species diversity within the confines of their permit (longer fishing seasons, broader spatial area). Using a long-term dataset of individual salmon fishers in the Alaska, we develop new models for understanding how the effects of diversification have changed through time. These changes allow us to compare patterns across regions and gear, as well as better understand how external forces have changed incentives for diversifying (hatchery production and price trends, EVOS?).

@@

Previous analyses have largely focused on patterns of diversification across ecosystems and the benefits to adopt more or less diverse fishing strategies [@kasperski_income_2013; @anderson_benefits]. Fisheries in Alaska have been studied extensively, because they are one of the most productive regions in North America (both by landed weight, and revenue [@office_of_science_and_technology_fisheries_2016]) and they allow individuals to adopt a wide range of fishing strategies. [@anderson_benefits] defined strategies as a permit or collection of permits held by single individuals. While adding new permits allows fishers to increase species diversity, few individual permits allow individuals to harvest more than one species (examples of single-species permits in the Gulf of Alaska include halibut, sablefish, herring, crab, shellfish, shrimp, sea cucumber). Some exceptions to these single species permits include participants in federally managed groundfish fisheries and participants in state managed salmon fisheries.

Understanding the effects of species diversification on salmon fishers is particularly important, as the number of salmon specialists (defined as fishers holding a single salmon permit) has increased over the last 30 years as a fraction of the total [@anderson_benefits]. For these limited entry salmon fisheries, general trends that have emerged are that more people are participating since the early 2000s (Fig. A1), median revenue is increasing (across individuals, Fig. A2), and landed species diversity declining over time (averaged across indivduals, Fig. A6). Across different salmon fisheries, other than a slight negative effect of diversification on variability (similar to [@anderson_benefits]), there appears to be little correlation between diversity and revenue (Fig. A4). In terrestrial agriculture, there are numerous examples where diversification may affect the mean or CV but not both [@purdy_financial_1997], and the same result may be true for salmon fishers. Further, these impacts of diversification may be variable over time.

Five species of salmon are encountered in Alaska waters (Chinook [*Oncorhynchus tshawytscha*], chum [*O. keta*], pink [*O. gorbuscha*], sockeye [*O. nerka*], coho [*O. kisutch*]), and the ability of individual salmon fishers to diversify across species is in part a function of geography and gears. For example, salmon may be caught with purse seine gear (n = 6 areas), drift gillnets (5 areas), and set gillnets (12 areas). Other gears include beach seine and troll gears (though these are generally restricted to Kodiak and Southeast Alaska, respectively). While some of these combinations of gears and areas allow fishers to diversify, there are many examples of permits where the catch is dominated by a single species. Perhaps the best examples include fisheries that target sockeye salmon (e.g. Bristol Bay set and drift gillnet fisheries). For these fisheries, specialization is associated with higher revenue and reduced CV in annual revenue [@anderson_benefits].

A secondary motivation for focusing on the effects of diversification for salmon fishers is to better understand the long term consequences of hatchery salmon production on the revenue and variability in revenue of individuals. Following low returns and declining productivity in the mid 1980s, many hatchery programs in the state rapidly increased hatchery releases [@mahnken_historical_1998; @leber_stock_2008]. Many of these hatchery programs increased production to stabilize the variability of fishers catch (and revenue), though these programs may also have costs to wild production [@ruggerone_productivity_2015]. The intra- and inter-specific effects of large scale hatchery programs has been debated [@hilborn_review_2001; @hilborn_hatcheries_1992; @wertheimer_comment:_2001; @heard_hatchery], but the effects of hatcheries have largely been focused on ecological interactions of aggregate yield to the fishery [@leber_stock_2008] or economic impacts [boyce_alaska_1993], rather than to variability experienced by individual fishermen. [Other history here?]

@@

In addition to the changing species composition of fish avaialble to be harvested, the benefits (or costs) of diversification are likely affected by a number of exogenous drivers, including marketing, prices, new processing technology [@knapp_trends]. [Break out as own paragraph+add more examples?] Following price crashes in the early 2000s, many of these factors have been positive for salmon fishermen (increasing trends in prices and catches, Fig. A3, A12). In contrast, the late 1980s - early 1990s experienced high abundance and low prices (in part due to collapses of foreign currencies). Combined with the increased trend in hatchery production and resulting changes in species compositon of adult salmon, these external factors may have also altered incentives for diversification. If there's an incentive to substitute high value species (Chinook, sockeye) with more abundant low value species (pink, chum), the analog in the terrestrial framring world would be crop substitution of low-volume high value crops (spices) with high-volume low value crops (cereals) (other examples in literature = cocoa, tobacco).

@@

We obtained fisheries landings and revenue data for all permit holders in Alaska from 1985 - 2014 from the Commercial Fisheries Entry Commission (CFEC). We adjusted for inflation by setting all revenues in 2009 USD. Because were were only interested in single permit fisheries, we adopted a simplified version of the filtering steps described in [@anderson_benefits]. After aggregating individual fish tickets to annual data, we retained individuals who only fished one of the six salmon permits included in a year (Table 1, Fig. A5). Like [@anderson_benefits] we also worked with a log-differenced dataset to remove non-stationary individuals; to avoid complications of individuals switching strategies, we only included pairs of years for a given person where they did not change strategies.

@@

Hatchery release trends for some regions included in our data have been summarized previously (e.g. [@ward_evaluating] summarized the trends for Prince William Sound). Release patterns have been dominated by pink and chum salmon (for PWS, these species represent 80% and 18% of total releases for 2012-2014 respectively, [@ward_evaluating]) and production has increased since the late 1970s (217 million juvenile pink salmon in PWS in 1985 versus 672 million in 2014). Though our analysis with CFEC data doesn't separate hatchery versus wild fish, as a contrast we can make coarse comparisons across regions and species to examine potential effects of hatchery fish on variability. For example, we can compare the variation in pink salmon returns (earnings or weight) from Prince William Sound (largely hatchery) to those from Southeast Alaska (largely wild). Similarly, we can compare the variability in earnings from species dominated by hatchery returns (pink, chum) to those with less hatchery influence (Chinook, sockeye). To make these comparisons, we aggregated the earnings and landings by species and permit, then calculated 3-year rolling CVs. As a second comparison, we examined whether long-term participants in any of the fisheries had experienced a change in variability. Similar to [@kasperski_income_2013] who calculated CVs of vessels before and after significant managment events, we identified individuals who fished the same permit for at least three years in the first decade of our data (1985-1995) and at least three years in the last decade of the data (2004-2014).

@@

One of the challenges in modeling the effects of species diversification for individual fishers is that between two years, both the species composition of the catch and the species composition of the fish available to be caught may change (making it hard to disentangle the selectivity of fishermen versus portfolio of fish available). To isolate the effects of a changing environment from changing strategies, we used a subset of the data from fishery permits where discrete strategies were emergent. For details see Appendix.

@@

Our summaries of changing variability (aggregate for each fishery, across individuals, or within individuals) was not consistent with the hypothesis that variability in revenue or landings has decreased over time (Fig. A9-A10). For pink and chum salmon, fisheries in PWS generally experience higher year to year variability compared to SEAK. In looking at the changing CV of individuals, we see a slight increase in the CV of revenue for people fishing in the recent decade who were also fishing 1985-1995 (with the exception being S01A).

@@

Figure 3: One component of variability is the year-to-year variation that individuals experience as a result of external forces (stock abundance, prices, etc). Or common fluctuations in revenue experienced by all permit holders [@anderson_benefits]. This shows the time varying intercepts on the mean -- negative equals losing revenue on average, positive = increasing revenue. This shows big differences between permit types (gillnets tend to have lower year-to-year variation). Some permits have extreme variation - S01E (PWS seine) has fluctuations between exp(-1) and exp(+1), corresponding to year-to-year changes of 36-270%.

Figure 4 shows the time varying nature of of the model for all permits. These are the time series of benefits of changing from effective species diversity of 1 to 2, and here we're basically isolating year-to-year differences in revenue (the intercepts) from diversification benefits. In some cases the effect is positive (fishing more spp = more revenue). But in many cases it's not, and the trend is negative (indicating that specialization is more profitable over time). In some cases, these time series mirror the aggregate diversity in the raw time series (Fig. 1). I think we also expect this result - fishermen increase diversity or not, because it's profitable.

@@

- Ability to specialize primarily driven by many forces beyond control of fishermen. But in some situations, diversification is possible by extending fishing season to exploit additional spp (fishing earlier for chum in PWS S01E). Or fishing different areas to exploit more spp in S03 (SEAK).

- Prices are generally synchronous for all spp (especially coho/chum/pink), and have increased at a faster rate compared to chum and pink salmons since 2000 (Fig A8-A9). We're also seeing a trend toward greater synchrony between species (Fig A11-A12, A14). Projecting future prices are complex [@knapp_trends]. Synchrony in prices may largely be function of marketing, improved product, freezing, etc.

@@

```{r, echo=FALSE, fig.pos="placeHere", fig.cap = "As model validation, we can compare the estimated intercepts for PWS S01E (purse seine) representing average year to year change in revenue versus the change year over year in ln(pink salmon returns). Correlation between them is ~ 0.8. This suggests (as expected) people on average make a lot more money in years when pink salmon returns are high."}

>>>>

---
title: 'Reduced diversity and increased specialization of Alaska salmon fishermen:
  effects on variation in individual revenues'
author: "Eric, Sean, Ole, Rich, Milo, Jordan, Jen, Alan, others"
date: ''
output:
  pdf_document:
    fig_caption: yes
  html_document: default
  word_document: default
bibliography: refs.bib
---

@@

On a large scale, individual fishers may experience positive outcomes from diversifying their catches; examples include reduced year-to-year variability in annual revenue and increases in average revenue. Fishers can increase their species diversity by owning a permit that allows multiple species to be targeted, or by owning multiple permits. External forces such as management regulations (limited entry,  catch shares) or market forces (permit or fish prices) may disincentive individuals to hold multiple permits. Thus, individuals with a single permit may increase species diversity within the confines of their permit (longer fishing seasons, broader spatial area). Using a long-term dataset of individual salmon fishers in Alaska, we develop new models for understanding how the effects of diversification have changed through time. These changes allow us to compare patterns across regions and gear, as well as better understand how external forces have changed incentives for diversifying (hatchery production and price trends, EVOS?).

@@

Previous analyses have largely focused on patterns of diversification across ecosystems and the benefits to adopt more or less diverse fishing strategies [@kasperski_income_2013; @anderson_benefits]. Fisheries in Alaska have been studied extensively, because they are one of the most productive seafood regions in the world (both by landed weight, and revenue [@office_of_science_and_technology_fisheries_2016], FAO cite) and they allow individuals to adopt a wide range of fishing strategies. [@anderson_benefits] defined strategies as a permit or collection of permits held by single individuals. While adding new permits allows fishers to increase species diversity, few individual permits allow individuals to harvest more than one species (examples of single-species permits in the Gulf of Alaska include halibut, sablefish, herring, crab, shellfish, shrimp, sea cucumber). Some exceptions to these single species permits include participants in federally managed groundfish fisheries and participants in state managed salmon fisheries.

Understanding the effects of species diversification on salmon fishers is particularly important, as the number of salmon specialists (defined as fishers holding a single salmon permit) has increased over the last 30 years as a fraction of the total [@anderson_benefits]. For these limited entry salmon fisheries, general trends that have emerged are that more people are participating since the early 2000s (Fig. A1), median revenue is increasing (across individuals, Fig. A2), and landed species diversity declining over time (averaged across indivduals, Fig. A6). Across different salmon fisheries, other than a slight negative effect of diversification on variability (similar to [@anderson_benefits]), there appears to be little correlation between diversity and revenue (Fig. A4). In terrestrial agriculture, there are numerous examples where diversification may affect the mean or CV but not both [@purdy_financial_1997], and the same result may be true for salmon fishers. Further, these impacts of diversification are variable over time.

Five species of salmon are encountered in Alaska waters (Chinook [*Oncorhynchus tshawytscha*], chum [*O. keta*], pink [*O. gorbuscha*], sockeye [*O. nerka*], coho [*O. kisutch*]), and the ability of individual salmon fishers to diversify across species is in part a function of geography and fishing gears. For example, salmon may be caught with purse seine gear (n = 6 areas), drift gillnets (5 areas), and set gillnets (12 areas). Other gears include beach seine and troll gears (though these are generally restricted to Kodiak and Southeast Alaska, respectively). While some of these combinations of gears and areas allow fishers to diversify, there are many examples of permits where the catch is dominated by a single species. Perhaps the best examples include fisheries that target sockeye salmon (e.g. Bristol Bay set and drift gillnet fisheries). For these fisheries, specialization is associated with higher revenue and reduced CV in annual revenue [@anderson_benefits].

A secondary motivation for focusing on the effects of diversification for salmon fishers is to better understand the long term consequences of hatchery salmon production on the revenue and variability in revenue of individuals. Following low returns and declining productivity in the mid 1980s, many hatchery programs in the state rapidly increased hatchery releases [@mahnken_historical_1998; @leber_stock_2008]. Many of these hatchery programs increased production to stabilize the variability of fishers catch (and revenue), though these programs may also have impacts on wild production [@ruggerone_productivity_2015]. The intra- and inter-specific effects of large scale hatchery programs has been debated [@hilborn_review_2001; @hilborn_hatcheries_1992; @wertheimer_comment:_2001; @heard_hatchery], but the effects of hatcheries have largely been focused on ecological interactions of aggregate yield to the fishery [@leber_stock_2008] or economic impacts [boyce_alaska_1993], rather than to variability experienced by individual fishermen. [Other history here?]

@@

In addition to the changing species composition of fish avaialble to be harvested, the benefits (or costs) of diversification are affected by a number of exogenous drivers, including marketing, prices, new processing technology [@knapp_trends]. [Break out as own paragraph+add more examples?] Following price crashes in the early 2000s, many of these factors have been positive for salmon fishermen (increasing trends in prices and catches, Fig. A3, A12). In contrast, the late 1980s - early 1990s experienced high abundance and low prices (in part due to collapses of foreign currencies). Combined with the increased trend in hatchery production and resulting changes in species compositon of adult salmon, these external factors may have also altered incentives for diversification. If there's an incentive to substitute high value species (Chinook, sockeye) with more abundant low value species (pink, chum), the analog in the terrestrial framring world would be crop substitution of low-volume high value crops (spices) with high-volume low value crops (cereals) (other examples in literature = cocoa, tobacco).

@@

We obtained fisheries landings and revenue data for all permit holders in Alaska from 1985 - 2014 from the Commercial Fisheries Entry Commission (CFEC). We adjusted for inflation by setting all revenues in 2009 USD. Because we were only interested in single permit fisheries, we adopted a simplified version of the filtering steps described in [@anderson_benefits]. After aggregating individual trip-level fish tickets to annual data, we retained individuals who only fished one of the six salmon permits included in a year (Table 1, Fig. A5). Like [@anderson_benefits] we also worked with a log-differenced dataset to remove non-stationary individuals; to avoid complications of individuals switching strategies, we only included pairs of years for a given person where they did not change strategies.

@@

Hatchery release trends for some regions included in our data have been summarized previously (e.g. [@ward_evaluating] summarized the trends for Prince William Sound). Release patterns have been dominated by pink and chum salmon (in recent years 2012-2014 in PWS, for example, pink and chum salmon represent 80% and 18% of total releases, [@ward_evaluating]) and production has increased since the late 1970s (217 million juvenile pink salmon in PWS in 1985 versus 672 million in 2014). Though our analysis with CFEC data doesn't separate hatchery versus wild fish, as a contrast we can make coarse comparisons across regions and species to examine potential effects of hatchery fish on variability. For example, we can compare the variation in the largely hatchery pink salmon returns (earnings and weight) from Prince William Sound to the largely wild pink salmon returns from Southeast Alaska. Similarly, we can compare the variability in earnings from species dominated by hatchery returns (pink, chum) to those with less hatchery influence (Chinook, sockeye). To make these comparisons, we aggregated the earnings and landings by species and permit, then calculated 3-year rolling CVs. As a second comparison, we examined whether long-term participants in any of the fisheries had experienced a change in variability. Similar to [@kasperski_income_2013] who calculated CVs of vessels before and after significant managment events, we identified individuals who fished the same permit for at least three years in the first decade of our data (1985-1995) and at least three years in the last decade of the data (2004-2014).

@@

One of the challenges in modeling the effects of species diversification for individual fishers is that between two years, both the species composition of the catch and the species composition of the fish available to be caught change (making it hard to disentangle the selectivity of fishermen versus portfolio of fish available). To isolate the effects of a changing environment from changing strategies, we used a subset of the data from fishery permits where discrete strategies were emergent. For details see Appendix.

@@

Our summaries of changing variability (aggregate for each fishery, across individuals, or within individuals) was not consistent with the hypothesis that variability in revenue or landings has decreased over time (Fig. A9-A10). For pink and chum salmon, fisheries in PWS generally experience higher year-to-year variability compared to SEAK. In looking at the changing CV of individuals, we see a slight increase in the CV of revenue for people fishing in the recent decade who were also fishing 1985-1995 (with the exception being S01A).

@@

Figure 3 shows that one component of variability is the year-to-year variation that individuals experience as a result of external forces (stock abundance, prices, etc) as wekk as common fluctuations in revenue experienced by all permit holders (Anderson et al., n.d.). This shows the time varying intercepts on the mean -- negative equals losing revenue on average, while positive implies increasing revenue. The panels of this figure show big differences between permit types (gillnets tend to have lower year-to-year variation). Some permits have extreme variation; for example, S01E (PWS seine) has fluctuations between exp(-1) and exp(+1), corresponding to year-to-year changes of 36-270%.

Figure 4 shows the time varying nature of of the model for all permits. These are the time series of benefits of changing from effective species diversity of 1 to 2, and here we are isolating year-to-year differences in revenue (the intercepts) from diversification benefits. In some cases the effect is positive (fishing more spp yields more revenue). But in many cases this is not the case, and the trend is negative (indicating that specialization is more profitable over time). In some cases, these time series mirror the aggregate diversity in the raw time series (Fig. 1). I think we also expect this result - fishermen increase diversity or not, because it's profitable.

@@

- Ability to specialize primarily driven by many forces beyond control of fishermen. But in some situations, diversification is possible by extending the fishing season to exploit additional spp (fishing earlier for chum in PWS S01E). Or fishing different areas to exploit more spp in S03 (SEAK).

- Prices are generally synchronous for all spp (especially coho/chum/pink), and have increased at a faster rate compared to chum and pink salmons since 2000 (Fig A8-A9). We're also seeing a trend toward greater synchrony between species (Fig A11-A12, A14). Projecting future prices is notably complex [@knapp_trends]. Synchrony in prices may largely be function of marketing, improved product, freezing, etc.

@@

```{r, echo=FALSE, fig.pos="placeHere", fig.cap = "As model validation, we can compare the estimated intercepts for PWS S01E (purse seine) representing average year to year change in revenue versus the change year over year in ln(pink salmon returns). Correlation between them is ~ 0.8. This suggests (as expected) people on average make a lot more money in years when pink salmon returns are high. [side note: diversification may only be beneficial here when returns and prices are both low.]"}
